Very simplified likelihood

### 2023-11-24 inst/very_simplified_likelihood.Rmd
more work including a vs b a/(a+b) horizontal axis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

@@

```{r}
 a <- seq(from=0.05,to=0.45,by=0.01)
 L <- f(a,r=c(3,4,3,3,1,3,3,3,1,1,1,5),10,log=T)  # 10+1==11 competitors in total
plot(a,L-max(L))
abline(v=a[which.max(L)])
abline(v=1/11)
abline(h=c(0,-2))
```

@@

plot(NA,xlim=c(0,1),ylim=c(-10,0),type='n')
abline(v=1/23)
for(i in seq_len(9)){points(a,fun(i),type='b')}

plot(NA,xlim=c(-5,1),ylim=c(-5,0),type='n',xlab="log-odds")
abline(v=-log(22))
for(i in seq_len(23)){points(log(a/(1-a)),fun(i),type='l')}
```

@@

plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
abline(v=-log(22))
for(i in seq_len(20)){points(log(a/(1-a)),fun(i),type='l')}

@@

```{r}
x <- as.matrix(read.table("olympiad.txt",header=TRUE))
x <- x[apply(x,1,max) <= 100,]
fun <- function(i){
 jj <- as.numeric(unlist(x[i,]))
  L <- f(a,jj,100,log=TRUE)
  return(L - max(L))
}
plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
abline(v=-log(99))
for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}

```{r}
x <- as.matrix(read.table("ol.txt",header=TRUE))[,1:4]
x <- x[apply(x,1,max) <= 100,]
fun <- function(i){
 jj <- as.numeric(unlist(x[i,]))
  L <- f(a,jj,100,log=TRUE)
  return(L - max(L))
}
plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
abline(v=-log(99))
for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}
```

```{r}
x <- as.matrix(read.table("ol.txt",header=TRUE))[,1:4]
x <- x[apply(x,1,max) <= 100,]
x[] <- sample(c(x))
x
rowSums(x)
fun <- function(i){
 jj <- as.numeric(unlist(x[i,]))
  L <- f(a,jj,100,log=TRUE)
  return(L - max(L))
}
plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
abline(v=-log(99))
for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("pracma")
```

@@

```{r}
 a <- seq(from=0.05,to=0.45,by=0.01)
 L <- f(a,r=5,10,log=T)  # 10+1==11 competitors in total
plot(a,L-max(L))
abline(v=a[which.max(L)])
abline(v=1/11)
abline(h=c(0,-2))
```

@@

plot(NA,xlim=c(0,1),ylim=c(-10,0),type='n')
for(i in seq_len(9)){points(a,fun(i),type='b')}

plot(NA,xlim=c(-3,3),ylim=c(-5,0),type='n',xlab="log-odds")
for(i in seq_len(23)){points(log(22*a/(1-a)),fun(i),type='l')}
```

@@

plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
for(i in seq_len(20)){points(log(22*a/(1-a)),fun(i),type='l')}

@@

```{r}
n <- 82 # 83 competitors altogether
x <- as.matrix(read.table("olympiad.txt",header=TRUE))
x <- x[apply(x,1,max) <= 100,]
fun <- function(i){
 jj <- as.numeric(unlist(x[i,]))
  L <- f(a,jj,100,log=TRUE)
  return(L - max(L))
}

plot(NA,xlim=c(-5,5),ylim=c(-5,1),type='n',xlab="log-odds")
for(i in seq_len(40)){points(log(n*a/(1-a)),fun(i),type='l')}
abline(v=0)
```

```{r}
x <- as.matrix(read.table("olympiad.txt",header=TRUE))[,1:4]
x <- x[apply(x,1,max) <= 100,]
fun <- function(i){
 jj <- as.numeric(unlist(x[i,]))
  L <- f(a,jj,100,log=TRUE)
  return(L - max(L))
}
plot(NA,xlim=c(-5,5),ylim=c(-5,1),type='n',xlab="log-odds")
for(i in seq_len(40)){points(log(n*a/(1-a)),fun(i),type='l')}
```

```{r}
x <- as.matrix(read.table("olympiad.txt",header=TRUE))[,1:4]
x <- x[apply(x,1,max) <= 100,]
x[] <- sample(c(x))
head(x)
rowSums(x)
fun <- function(i){
 jj <- as.numeric(unlist(x[i,]))
  L <- f(a,jj,100,log=TRUE)
  return(L - max(L))
}
plot(NA,xlim=c(-5,5),ylim=c(-5,1),type='n',xlab="log-odds")
for(i in seq_len(40)){points(log(n*a/(1-a)),fun(i),type='l')}
```

In principle we can find the optimum point with calculus and indeed we
have

\[
\frac{d}{db}\mathcal{L} = \frac{\psi(1-1/b)-\psi(r+1-1/b)}{b(b+\psi(1-1/b)-\psi(r+1-1/b))}-n
\]

but the implementation is difficult because the expression is not
defined for many sensible values of $b$.  We can use `optimize()`
though:

```{r solvescore}
  MLE <- function(r,n){
  small <- 1e-6
  a <- optimize(function(b){
       f_single_direct(b,r=r,n=n,log=TRUE)},
       maximum=TRUE, c(small,1-small))$maximum
  b <- (1-a)/n
  print(c(a,b))
  return(a/(a+b))
}

MLE(0,10)
MLE(1,10)
MLE(2,10)
MLE(3,10)
MLE(4,10)
MLE(5,10)
MLE(6,10)

MLE(0,11)
MLE(1,11)
MLE(2,11)
MLE(3,11)
MLE(4,11)
MLE(5,11)
MLE(6,11)
